Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-13 index.qmd
section:background,structural; process:write,figure

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the method's outcome. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. For instance, @vanDaal_et_al_2016 assessed academic writing skills of university students (units) using multiple argumentative essays (sub-units).

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021; @Bouwer_et_al_2023], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

@@

This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs. The section is not a comprehensive description of causal inference methods. Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020. They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful. For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.

## The empirical research endeavor {#sec-framework-empirical}

<!-- 1. What is the purpose of the research endeavor? -->
The goal of an investigator is to generate new knowledge or deepen the understanding of specific topics or issues. Investigators accomplish this by addressing research problems. Research problems fall into two categories: empirical and non-empirical research. The first tackles real-world challenges, relying on evidence gathered through observation and experimentation. The latter focuses on conceptual inquiries that enhance understanding complex concepts, theories, and phenomena.

<!-- 2. How to approach a research problem?  -->
<!-- 2.1 the role of a research question -->
Addressing research problems requires formulating well-defined research questions that provide focus, clarity, and structure to a study. In empirical research, these questions are often framed as *targets of inference* or *estimands*, representing the quantities the investigator aims to estimate. For instance, an investigator might want to examine how the duration of cochlear implant use affects the average overall speech quality of a child.

<!-- 2.2 what else do you need? -->
Furthermore, in empirical research, investigators must design appropriate methodologies, collect data, and ensure a systematic and coherent analysis of the results. To achieve this, they must accurately identify the factors influencing an outcome and correctly attribute the effects to each factor. Consequently, understanding how an investigator transitions from a target of inference to generating statistical estimates is crucial.

<!-- 3. So, how do we move from targets of inference to estimates -->
@fig-IEflow shows the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], which illustrates the mechanism for transitioning from a estimand to a corresponding *statistical estimate*, through identification and estimation.

<!-- Specifically, the figure an investigator must define the different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method. The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data. Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets. Bayesian statistical methods offer an intuitive approach for the task. Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands. Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis. -->

<!-- translate the encoded causal information into a probabilistic statistical model. Then, using this model along with data, the investigator must produce an estimate that addresses the research problem of interest. This procedure is illustrated in the Identification-Estimation flowchart presented in @fig-IEflow. -->

<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*. -->

![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}

@@

<!-- 1. What is causal inference? -->
<!-- Causal inference is essential to the empirical research endeavor, as often, investigators want to make causal claims, rather than merely associational claims.  -->

In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical notation for analyzing causes and counterfactuals [@Pearl_2009].

<!-- 2. what are counterfactuals? -->
Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."

<!-- 3. How do we apply causal inference and counterfactuals? -->
Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].

<!-- 4. But which approach are we going to use? -->
However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016]. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].

@@

<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=90%}

>>>>

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the method's outcome. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. For example, in @vanDaal_et_al_2016, the authors assessed the academic writing skills of university students (units) using multiple argumentative essays (sub-units).

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021; @Bouwer_et_al_2023], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach to causal inference, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

@@

This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs. The section is not a comprehensive description of causal inference methods. Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020. They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful. For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books like @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.

@@

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might seek to answer the question, "To what extent do daily practice hours ($X$) influence students' ability to produce high-quality written texts ($Y$)?". To investigate this, the study randomly assigns students to two groups with different levels of daily practice, low $(X=1)$ and medium $(X=2)$. The quality of the written texts is evaluated using pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$. The research question is then framed as the estimand, "*On average*, which group of students produces higher-quality written texts?" This estimand is mathematically expressed as $E[Y|X=1] - E[Y|X=2]$, where $E[\cdot]$ represents the expected value.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* is the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010]. While various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure accurate estimates. For example, Fisher's exact test [@Fisher_1922] is an estimator with desirable properties for comparing $2\times2$ tables. The test provides accurate estimates for comparing proportions when its assumptions are satisfied. @fig-IEflow, known as the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], illustrates the aforementioned process of transitioning from estimands to estimates.

![ Identification-Estimation flowchart. Extracted and slightly modified from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}

<!-- 3. But we want these estimands to be causal -->
However, many studies aim to establish causal relationships rather than merely infer associations. In the earlier example, the estimates obtained using Fisher's exact test can be interpreted as causal because the data were collected through a randomized experiment. In randomized experiments, the researcher fully controls the treatment assignment mechanism, such as assigning students to different groups. This control allows causal estimates to be derived from associational estimates, such as those provided by the Fisher statistic, since randomization effectively eliminates confounding [@Neal_2020]. *Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious association [@Everitt_et_al_2010].

<!-- 4. causal inference to the rescue -->
<!-- Nevertheless, in many cases, researchers are not able to perform experiments due to many reasons. For instance, . In this scenarios, the framework of causal inference can be an alternative to generate causal estimates. *Causal inference* is a framework focused on identifying the causes influencing a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which primarily focuses on summarizing data and inferring associations, causal inference provides a coherent mathematical framework for analyzing causes and counterfactuals [@Pearl_2009]. -->

<!-- 5. How does causal inference defines causal effects? what are counterfactuals? -->
<!-- Causal inference defines causal effects in terms of counterfactuals. Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. Returning to the previous example, if the research aim to determine the $X$ have an impact on an outcome $Y$ -->

<!-- Causal inference approach the *fundamental problem of causal inference* -->

<!-- According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances." -->

<!-- 6. How do we apply causal inference and counterfactuals? -->
<!-- Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010]. -->

<!-- 7. But which approach are we going to use? -->
<!-- However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016]. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. -->

@@

<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=80%}
